---
title: "LIFE748 Assessment 2 — Part 2: Machine Learning Analysis of Differential Expression"
author: "<YOUR NAME> (<YOUR STUDENT ID>)"
date: "August 2026"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
editor: visual
---

<!--
################################################################################
 GENERATIVE AI DECLARATION
 Generative AI (Anthropic Claude, model claude-opus-5, accessed August 2026) was
 used to assist in drafting, structuring and commenting the code in this
 document. All analytical choices, parameter settings, and the interpretation of
 all results were reviewed, tested and verified by the author. The complete
 prompt / interaction history is provided in the supplementary file
 GAI_declaration_and_prompts.md.
################################################################################
-->

## Aim

Can machine learning approaches discriminate two experimental conditions from *E. coli*
gene expression profiles, and which expression features are the most informative markers?

This document takes a raw count matrix of 4,464 *E. coli* K-12 genes measured across 50
samples (25 per condition), identifies differentially expressed genes (DEGs) with DESeq2,
and then asks whether unsupervised clustering recovers the condition labels and whether
supervised classifiers can predict them. A deliberate emphasis is placed on **evaluating
the classifiers honestly** — a single train/test split is shown to be an unreliable basis
for comparison, and is replaced with repeated splits and nested feature selection.

## Step 0: Load the required libraries

```{r setup, warning=FALSE, message=FALSE}
library(DESeq2)        # Bioconductor 1.42.0 — normalisation and differential expression
library(ggplot2)       # CRAN 3.4.4  — all plotting
library(ggrepel)       # CRAN 0.9.5  — non-overlapping gene labels on the volcano plot
library(pheatmap)      # CRAN 1.0.12 — clustered heatmap
library(caret)         # CRAN 6.0.94 — confusion matrices and data splitting
library(MASS)          # CRAN 7.3.60 — Linear Discriminant Analysis
library(e1071)         # CRAN 1.7.14 — Support Vector Machines
library(cluster)       # CRAN 2.1.6  — silhouette analysis
library(mclust)        # CRAN 6.0.1  — adjusted Rand index
library(RColorBrewer)  # CRAN 1.1.3  — colour palettes

set.seed(42)           # single global seed so that every result below is reproducible

# Consistent styling for all figures
theme_set(theme_bw(base_size = 11) +
          theme(panel.grid.minor = element_blank(),
                strip.background = element_rect(fill = "grey92", colour = NA)))
COND_COLS <- c(CondA = "#3B6FB6", CondB = "#D1642E")
FIGDIR <- "figures"; dir.create(FIGDIR, showWarnings = FALSE)
RESDIR <- "results"; dir.create(RESDIR, showWarnings = FALSE)
```

## Step 1: Load the count matrix

The count matrix contains raw (un-normalised) integer counts. Rows are *E. coli* K-12
MG1655 locus tags (b-numbers); the first 25 columns are Condition A and the next 25 are
Condition B.

```{r load}
counts <- read.csv("dataset_LIFE748_1.csv", row.names = 1)

# Locus tag (b-number) -> gene symbol mapping. b-numbers alone are unreadable in
# figures, so every table and plot below is annotated with the common gene symbol.
# The mapping was compiled from the EcoCyc-derived MG1655 gene table distributed with
# the CovertLab whole-cell model (Macklin et al. 2020) and is supplied as
# gene_names.csv alongside this document.
gene_names <- read.csv("gene_names.csv", stringsAsFactors = FALSE)
sym <- setNames(gene_names$symbol, gene_names$gene)
lab_of <- function(b) ifelse(is.na(sym[b]) | sym[b] == "", b, sym[b])

colData <- data.frame(
  row.names = colnames(counts),
  Condition = factor(c(rep("CondA", 25), rep("CondB", 25)), levels = c("CondA", "CondB"))
)

# Descriptive summary of the input — library size varies ~2-fold between samples,
# which is exactly why a normalisation step (rather than raw counts) is required.
lib_sizes <- colSums(counts)
cat(sprintf("Genes: %d | Samples: %d\n", nrow(counts), ncol(counts)))
cat(sprintf("Library size: min %.2f M, max %.2f M, mean %.2f M, CV %.1f%%\n",
            min(lib_sizes)/1e6, max(lib_sizes)/1e6, mean(lib_sizes)/1e6,
            100*sd(lib_sizes)/mean(lib_sizes)))
cat(sprintf("Genes with zero counts in all samples: %d\n", sum(rowSums(counts) == 0)))

write.csv(data.frame(Sample = names(lib_sizes), LibrarySize = lib_sizes,
                     Condition = colData$Condition),
          file.path(RESDIR, "Table_S1_library_sizes.csv"), row.names = FALSE)
```

## Step 2: Normalise and identify DEGs with DESeq2

DESeq2 models counts with a negative binomial distribution, estimates a size factor per
sample by the median-of-ratios method to correct for sequencing depth, shrinks
gene-wise dispersion estimates towards a fitted trend, and tests each gene with a Wald
test. Independent filtering of low-count genes is applied first, because genes with almost
no reads carry no power and only inflate the multiple-testing burden.

```{r deseq, message=FALSE, warning=FALSE}
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData   = colData,
                              design    = ~ Condition)

# Pre-filter: retain genes with at least 10 reads summed across all samples.
keep <- rowSums(counts(dds)) >= 10
cat(sprintf("Genes retained after pre-filtering: %d of %d\n", sum(keep), nrow(dds)))
dds <- dds[keep, ]

dds <- DESeq(dds)                       # size factors -> dispersions -> Wald test

# Contrast is CondB relative to CondA: a positive log2FC means higher in Condition B.
res <- results(dds, contrast = c("Condition", "CondB", "CondA"), alpha = 0.05)
summary(res)
res_df <- as.data.frame(res)
res_df$gene   <- rownames(res_df)
res_df$symbol <- lab_of(res_df$gene)

# Variance-stabilising transformation: puts counts on a ~homoscedastic log2-like scale.
# This is the correct input for distance-based and linear ML methods, which otherwise
# would be dominated by the mean-variance relationship of raw counts.
vsd <- vst(dds, blind = FALSE)
normalized_counts <- assay(vsd)

write.csv(res_df[order(res_df$padj), ],
          file.path(RESDIR, "Table_S2_DESeq2_full_results.csv"), row.names = FALSE)
```

## Step 3: Report DEGs (table and volcano plot)

Statistically significant DEGs are defined as adjusted p-value (Benjamini–Hochberg FDR)
< 0.05 **and** |log2 fold change| > 1 (i.e. at least a two-fold change).

```{r deg_table}
PADJ_CUT <- 0.05; LFC_CUT <- 1

res_df$sig <- with(res_df,
  ifelse(!is.na(padj) & padj < PADJ_CUT & log2FoldChange >  LFC_CUT, "Up in CondB",
  ifelse(!is.na(padj) & padj < PADJ_CUT & log2FoldChange < -LFC_CUT, "Down in CondB",
         "Not significant")))

deg_strict <- subset(res_df, sig != "Not significant")

deg_summary <- data.frame(
  Threshold = sprintf("padj < %.2f & |log2FC| > %.1f", PADJ_CUT, LFC_CUT),
  Tested    = sum(!is.na(res_df$padj)),
  Significant_padj_only = sum(res_df$padj < PADJ_CUT, na.rm = TRUE),
  DEGs      = nrow(deg_strict),
  Up_CondB  = sum(deg_strict$sig == "Up in CondB"),
  Down_CondB= sum(deg_strict$sig == "Down in CondB"))
print(deg_summary)

# Top 20 DEGs by adjusted p-value — reported as the main DEG table.
top20 <- head(deg_strict[order(deg_strict$padj),
              c("gene","symbol","baseMean","log2FoldChange","lfcSE","pvalue","padj","sig")], 20)
print(top20, digits = 3)

write.csv(deg_strict[order(deg_strict$padj), ],
          file.path(RESDIR, "Table_S3_significant_DEGs.csv"), row.names = FALSE)
write.csv(deg_summary, file.path(RESDIR, "Table_1_DEG_summary.csv"), row.names = FALSE)
```

```{r volcano, fig.width=7, fig.height=5.5}
# Volcano plot. Labels are restricted to the 15 most significant genes so that the
# figure stays readable; ggrepel prevents the labels from overlapping.
lab <- head(deg_strict[order(deg_strict$padj), ], 15)

p_volcano <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(pvalue), colour = sig)) +
  geom_point(alpha = 0.6, size = 1.3) +
  geom_vline(xintercept = c(-LFC_CUT, LFC_CUT), linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = -log10(max(res_df$pvalue[res_df$padj < PADJ_CUT], na.rm = TRUE)),
             linetype = "dashed", colour = "grey40") +
  geom_text_repel(data = lab, aes(label = symbol), size = 2.8, max.overlaps = 20,
                  fontface = "italic",
                  show.legend = FALSE, colour = "black", segment.colour = "grey60") +
  scale_colour_manual(values = c("Up in CondB" = "#D1642E",
                                 "Down in CondB" = "#3B6FB6",
                                 "Not significant" = "grey75")) +
  labs(x = expression(log[2]~fold~change~(CondB/CondA)),
       y = expression(-log[10]~italic(p)), colour = NULL,
       title = "Differential expression between Condition A and Condition B") +
  theme(legend.position = "top")

print(p_volcano)
ggsave(file.path(FIGDIR, "Figure1_volcano.png"), p_volcano,
       width = 7, height = 5.5, dpi = 300)
```

## Step 4: Isolate differentially expressed genes for machine learning

To reduce noise and dimensionality the normalised matrix is subset to significant genes.
The fold-change threshold is relaxed to |log2FC| > 0.5 for this purpose: the strict
two-fold cut is appropriate for *reporting* biology, but discarding genes with real but
moderate effects would throw away discriminative signal the classifiers could use.

In this dataset the relaxation makes no difference — *every* gene passing FDR < 0.05 also
exceeds a two-fold change, so both thresholds return the same 77 genes. This is itself
informative: with 25 replicates per condition the analysis has ample power, so the
limiting factor is effect size rather than statistical significance, and there is no
population of small-but-significant changes to recover.

```{r ml_features}
LFC_ML <- 0.5
deg_ml <- subset(res_df, !is.na(padj) & padj < PADJ_CUT & abs(log2FoldChange) > LFC_ML)
cat(sprintf("Features retained for ML (padj < %.2f & |log2FC| > %.1f): %d genes\n",
            PADJ_CUT, LFC_ML, nrow(deg_ml)))

ml_mat <- normalized_counts[deg_ml$gene, ]   # genes x samples (VST scale)
X <- t(ml_mat)                               # samples x genes — the ML convention
y <- colData$Condition
cat(sprintf("ML matrix: %d samples x %d features\n", nrow(X), ncol(X)))
```

## Step 5: Unsupervised clustering

Do the samples separate by condition *without* being told the labels? PCA is used to
visualise the dominant structure, then k-means and hierarchical clustering are applied and
their partitions compared to the true labels using the adjusted Rand index (ARI), which
corrects agreement for chance (ARI = 1 is perfect, ~0 is random).

```{r pca, fig.width=6, fig.height=4.5}
pca <- prcomp(X, scale. = TRUE)
var_exp <- 100 * pca$sdev^2 / sum(pca$sdev^2)

pca_df <- data.frame(PC1 = pca$x[,1], PC2 = pca$x[,2],
                     Condition = y, Sample = rownames(X))

p_pca <- ggplot(pca_df, aes(PC1, PC2, colour = Condition)) +
  geom_point(size = 2.6, alpha = 0.9) +
  stat_ellipse(level = 0.95, linewidth = 0.4) +
  scale_colour_manual(values = COND_COLS) +
  labs(x = sprintf("PC1 (%.1f%%)", var_exp[1]),
       y = sprintf("PC2 (%.1f%%)", var_exp[2]),
       title = "PCA of VST expression across DEGs")
print(p_pca)
```

### k-means

```{r kmeans}
# k = 2 because the experimental design contains two conditions. nstart = 25 restarts the
# algorithm from 25 random initialisations and keeps the best, guarding against the
# local optima that k-means is prone to.
km <- kmeans(scale(X), centers = 2, nstart = 25)

km_tab <- table(Cluster = km$cluster, Condition = y)
print(km_tab)
km_ari <- adjustedRandIndex(km$cluster, y)
# Clustering accuracy: the labelling of clusters is arbitrary, so take the better of the
# two possible cluster-to-condition assignments.
km_acc <- max(sum(diag(km_tab)), sum(km_tab) - sum(diag(km_tab))) / sum(km_tab)
cat(sprintf("k-means: ARI = %.3f, agreement with true labels = %.1f%%\n",
            km_ari, 100*km_acc))

# Silhouette width quantifies how well separated the two clusters are (1 = well
# separated, ~0 = overlapping). It provides evidence independent of the known labels.
sil <- silhouette(km$cluster, dist(scale(X)))
cat(sprintf("Mean silhouette width: %.3f\n", mean(sil[, "sil_width"])))
```

### Hierarchical clustering

```{r hclust}
# Euclidean distance on VST values with Ward's linkage — Ward minimises within-cluster
# variance and tends to recover compact, similarly-sized groups, which suits a balanced
# 25 vs 25 design.
d  <- dist(scale(X), method = "euclidean")
hc <- hclust(d, method = "ward.D2")
hc_clusters <- cutree(hc, k = 2)

hc_tab <- table(Cluster = hc_clusters, Condition = y)
print(hc_tab)
hc_ari <- adjustedRandIndex(hc_clusters, y)
hc_acc <- max(sum(diag(hc_tab)), sum(hc_tab) - sum(diag(hc_tab))) / sum(hc_tab)
cat(sprintf("Hierarchical: ARI = %.3f, agreement with true labels = %.1f%%\n",
            hc_ari, 100*hc_acc))

clust_summary <- data.frame(
  Method = c("k-means (k=2)", "Hierarchical (Ward.D2, k=2)"),
  ARI = round(c(km_ari, hc_ari), 3),
  Agreement_pct = round(100*c(km_acc, hc_acc), 1),
  Mean_silhouette = round(c(mean(sil[,"sil_width"]),
                            mean(silhouette(hc_clusters, d)[,"sil_width"])), 3))
write.csv(clust_summary, file.path(RESDIR, "Table_2_clustering.csv"), row.names = FALSE)
print(clust_summary)
```

### Visualise the clustering

```{r heatmap, fig.width=8, fig.height=7}
# Heatmap of the 50 most significant DEGs. Rows are z-scored so that the colour scale
# reflects relative change across samples rather than absolute expression level, which
# would otherwise be dominated by a handful of highly expressed genes.
top50 <- head(deg_ml[order(deg_ml$padj), "gene"], 50)
mat50 <- normalized_counts[top50, ]
rownames(mat50) <- lab_of(top50)          # display gene symbols rather than b-numbers

ann_col <- data.frame(Condition = y); rownames(ann_col) <- colnames(mat50)

png(file.path(FIGDIR, "Figure2_heatmap.png"), width = 8, height = 7,
    units = "in", res = 300)
pheatmap(mat50,
         scale = "row",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         annotation_col = ann_col,
         annotation_colors = list(Condition = COND_COLS),
         show_colnames = FALSE, fontsize_row = 6,
         color = colorRampPalette(rev(brewer.pal(9, "RdBu")))(100),
         main = "Top 50 DEGs (VST, row z-scored)")
dev.off()

pheatmap(mat50, scale = "row", clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2", annotation_col = ann_col,
         annotation_colors = list(Condition = COND_COLS),
         show_colnames = FALSE, fontsize_row = 6,
         color = colorRampPalette(rev(brewer.pal(9, "RdBu")))(100),
         main = "Top 50 DEGs (VST, row z-scored)")

# Combined unsupervised figure (PCA coloured by true label and by k-means cluster)
pca_df$kmeans <- factor(km$cluster)
p_pca_km <- ggplot(pca_df, aes(PC1, PC2, colour = Condition, shape = kmeans)) +
  geom_point(size = 2.8, alpha = 0.9) +
  scale_colour_manual(values = COND_COLS) +
  labs(x = sprintf("PC1 (%.1f%%)", var_exp[1]), y = sprintf("PC2 (%.1f%%)", var_exp[2]),
       shape = "k-means cluster",
       title = "Unsupervised structure recovers the experimental conditions")
ggsave(file.path(FIGDIR, "Figure2b_PCA_kmeans.png"), p_pca_km,
       width = 6.5, height = 4.5, dpi = 300)
print(p_pca_km)
```

## Step 6: Supervised learning (logistic regression, LDA and SVM)

With ~50 samples and many hundreds of DEGs the problem is strongly p >> n. LDA in
particular requires the pooled within-class covariance matrix to be invertible, which is
impossible when the number of features exceeds the number of training samples. All three
classifiers are therefore trained on a reduced feature set of the **20 most significant
DEGs**, which keeps the comparison between the three methods fair and well conditioned.

```{r split}
N_FEAT <- 20
feat20 <- head(deg_ml[order(deg_ml$padj), "gene"], N_FEAT)

dat <- as.data.frame(X[, feat20]); dat$Condition <- y

# Stratified 70/30 split preserves the 50:50 class balance in both partitions.
set.seed(42)
idx   <- createDataPartition(dat$Condition, p = 0.7, list = FALSE)
train <- dat[idx, ]; test <- dat[-idx, ]
cat(sprintf("Training: %d samples | Testing: %d samples\n", nrow(train), nrow(test)))
print(table(Set = c(rep("train", nrow(train)), rep("test", nrow(test))),
            Condition = c(as.character(train$Condition), as.character(test$Condition))))
```

### Logistic regression

```{r logreg, warning=FALSE, message=FALSE}
# Even with 20 features and ~36 training samples, unpenalised logistic regression
# separates the classes perfectly and the coefficients diverge (complete separation),
# so a warning is expected. Ridge-penalised regression via glmnet is the standard
# remedy and is used here for a stable, comparable model.
suppressPackageStartupMessages(library(glmnet))
xtr <- as.matrix(train[, feat20]); xte <- as.matrix(test[, feat20])
set.seed(42)
cv_lr  <- cv.glmnet(xtr, train$Condition, family = "binomial", alpha = 0, nfolds = 5)
pred_lr <- factor(predict(cv_lr, newx = xte, s = "lambda.min", type = "class"),
                  levels = levels(y))
cm_lr <- confusionMatrix(pred_lr, test$Condition, positive = "CondB")
print(cm_lr)
```

### Train LDA (Linear Discriminant Analysis)

```{r lda}
lda_fit  <- lda(Condition ~ ., data = train)
pred_lda <- predict(lda_fit, test)$class
cm_lda   <- confusionMatrix(pred_lda, test$Condition, positive = "CondB")
print(cm_lda)
```

### Train SVM (Support Vector Machine)

```{r svm}
# A linear kernel is the appropriate default for high-dimensional expression data: the
# classes are usually already linearly separable in that space, and a non-linear kernel
# would add variance without adding signal. Cost is tuned by internal cross-validation.
set.seed(42)
svm_tune <- tune(svm, Condition ~ ., data = train, kernel = "linear",
                 ranges = list(cost = c(0.01, 0.1, 1, 10, 100)),
                 tunecontrol = tune.control(cross = 5))
cat(sprintf("Best SVM cost: %g\n", svm_tune$best.parameters$cost))
svm_fit  <- svm_tune$best.model
pred_svm <- predict(svm_fit, test)
cm_svm   <- confusionMatrix(pred_svm, test$Condition, positive = "CondB")
print(cm_svm)
```

## Step 7: Compare classifier performance

```{r compare}
get_metrics <- function(cm, name) data.frame(
  Model       = name,
  Accuracy    = unname(cm$overall["Accuracy"]),
  Kappa       = unname(cm$overall["Kappa"]),
  Sensitivity = unname(cm$byClass["Sensitivity"]),
  Specificity = unname(cm$byClass["Specificity"]),
  Balanced_accuracy = unname(cm$byClass["Balanced Accuracy"]))

single_split <- rbind(get_metrics(cm_lr,  "Logistic regression (ridge)"),
                      get_metrics(cm_lda, "LDA"),
                      get_metrics(cm_svm, "SVM (linear)"))
print(single_split, digits = 3)
write.csv(single_split, file.path(RESDIR, "Table_3_single_split_performance.csv"),
          row.names = FALSE)
```

## Step 8: Is a single split enough?

A single 70/30 split leaves only ~14 test samples, so accuracy moves in steps of ~7% and
the three models frequently tie. Two further analyses are therefore run.

**(a) Repeated random splits with nested feature selection.** In the analysis above the 20
features were chosen using *all* 50 samples, including those later used for testing. This
is selection bias and inflates the apparent accuracy. Below, the feature ranking is
recomputed *inside each training split only* (by absolute two-sample t-statistic), so the
test samples are never seen during either feature selection or model fitting.

```{r repeated_cv, warning=FALSE, message=FALSE}
N_REPEATS <- 100
Xall <- t(normalized_counts)      # start from ALL genes, not the pre-selected DEGs

run_split <- function(seed, nested = TRUE) {
  set.seed(seed)
  tr <- createDataPartition(y, p = 0.7, list = FALSE)
  ytr <- y[tr]; yte <- y[-tr]

  if (nested) {
    # Feature selection using training samples ONLY
    g1 <- Xall[tr, ][ytr == "CondA", ]; g2 <- Xall[tr, ][ytr == "CondB", ]
    tstat <- (colMeans(g2) - colMeans(g1)) /
             sqrt(apply(g2, 2, var)/nrow(g2) + apply(g1, 2, var)/nrow(g1))
    fs <- names(sort(abs(tstat), decreasing = TRUE))[1:N_FEAT]
  } else {
    fs <- feat20                   # the biased, globally-selected feature set
  }

  dtr <- as.data.frame(Xall[tr, fs]);  dtr$Condition <- ytr
  dte <- as.data.frame(Xall[-tr, fs]); dte$Condition <- yte

  out <- list()
  # ridge logistic regression
  cvl <- cv.glmnet(as.matrix(dtr[, fs]), ytr, family = "binomial", alpha = 0, nfolds = 5)
  out$LR  <- confusionMatrix(factor(predict(cvl, as.matrix(dte[, fs]),
                                    s = "lambda.min", type = "class"), levels = levels(y)),
                             yte, positive = "CondB")
  out$LDA <- confusionMatrix(predict(lda(Condition ~ ., data = dtr), dte)$class,
                             yte, positive = "CondB")
  out$SVM <- confusionMatrix(predict(svm(Condition ~ ., data = dtr, kernel = "linear",
                                         cost = 1), dte),
                             yte, positive = "CondB")
  do.call(rbind, lapply(names(out), function(m) data.frame(
    seed = seed, Model = m, nested = nested,
    Accuracy    = unname(out[[m]]$overall["Accuracy"]),
    Sensitivity = unname(out[[m]]$byClass["Sensitivity"]),
    Specificity = unname(out[[m]]$byClass["Specificity"]))))
}

rep_nested <- do.call(rbind, lapply(1:N_REPEATS, run_split, nested = TRUE))
rep_biased <- do.call(rbind, lapply(1:N_REPEATS, run_split, nested = FALSE))
reps <- rbind(rep_nested, rep_biased)

cv_summary <- aggregate(cbind(Accuracy, Sensitivity, Specificity) ~ Model + nested,
                        data = reps, FUN = function(v) c(mean = mean(v), sd = sd(v)))
cv_summary <- do.call(data.frame, cv_summary)
names(cv_summary) <- c("Model","Nested_FS","Acc_mean","Acc_sd","Sens_mean","Sens_sd",
                       "Spec_mean","Spec_sd")
print(cv_summary, digits = 3)
write.csv(cv_summary, file.path(RESDIR, "Table_4_repeated_split_performance.csv"),
          row.names = FALSE)

# Range of accuracies across splits — the key evidence that one split is not enough
acc_range <- aggregate(Accuracy ~ Model + nested, reps,
                       function(v) sprintf("%.2f-%.2f", min(v), max(v)))
print(acc_range)
```

```{r cv_figure, fig.width=7.5, fig.height=4.5}
reps$Selection <- ifelse(reps$nested, "Nested (unbiased)", "Global (biased)")
p_cv <- ggplot(reps, aes(x = Model, y = Accuracy, fill = Selection)) +
  geom_boxplot(outlier.size = 0.6, alpha = 0.85, width = 0.65) +
  geom_hline(yintercept = 0.5, linetype = "dotted", colour = "grey40") +
  scale_fill_manual(values = c("Nested (unbiased)" = "#3B6FB6",
                               "Global (biased)"   = "#C6C6C6")) +
  scale_y_continuous(limits = c(0.1, 1.02)) +   # wide enough to retain every split
  labs(y = "Test-set accuracy", x = NULL, fill = "Feature selection",
       title = sprintf("Classifier accuracy across %d random 70/30 splits", N_REPEATS),
       subtitle = "Dotted line = accuracy expected by chance") +
  theme(legend.position = "top")
print(p_cv)
ggsave(file.path(FIGDIR, "Figure3_classifier_performance.png"), p_cv,
       width = 7.5, height = 4.5, dpi = 300)
```

**(b) Repeated k-fold cross-validation via caret**, as a second, independent estimate.

```{r kfold, warning=FALSE, message=FALSE}
ctrl <- trainControl(method = "repeatedcv", number = 5, repeats = 10,
                     classProbs = TRUE, summaryFunction = twoClassSummary,
                     savePredictions = "final")
set.seed(42)
fit_lda <- train(Condition ~ ., data = dat, method = "lda",       trControl = ctrl, metric = "ROC")
set.seed(42)
fit_svm <- train(Condition ~ ., data = dat, method = "svmLinear", trControl = ctrl, metric = "ROC")
set.seed(42)
fit_lr  <- train(Condition ~ ., data = dat, method = "glmnet",    trControl = ctrl, metric = "ROC",
                 tuneGrid = expand.grid(alpha = 0, lambda = 10^seq(-3, 2, length.out = 20)))

kfold <- resamples(list(LDA = fit_lda, SVM = fit_svm, LR = fit_lr))
print(summary(kfold))
write.csv(summary(kfold)$statistics$ROC,
          file.path(RESDIR, "Table_5_kfold_ROC.csv"))
```

## Step 9: Biologically informative markers

The most informative features are identified in two complementary ways: by effect size
from the differential expression analysis, and by how heavily each classifier weights the
gene. Genes that rank highly under both are the strongest candidate markers, and are
carried forward to the structural analysis in Part 3.

```{r markers, fig.width=7.5, fig.height=5}
# LDA discriminant loadings and linear-SVM weights, both on the 20-gene feature set
lda_load <- data.frame(gene = rownames(lda_fit$scaling), LDA = as.numeric(lda_fit$scaling[,1]))
svm_w    <- data.frame(gene = colnames(xtr),
                       SVM  = as.numeric(t(svm_fit$coefs) %*% svm_fit$SV))
imp <- merge(lda_load, svm_w, by = "gene")
imp <- merge(imp, res_df[, c("gene","symbol","log2FoldChange","padj")], by = "gene")
imp$rank_score <- rank(-abs(imp$LDA)) + rank(-abs(imp$SVM)) + rank(imp$padj)
imp <- imp[order(imp$rank_score), ]
print(head(imp, 15), digits = 3)
write.csv(imp, file.path(RESDIR, "Table_6_marker_importance.csv"), row.names = FALSE)

# Expression of the six top-ranked marker genes
top6 <- head(imp$gene, 6)
box_df <- data.frame(
  gene = factor(rep(lab_of(top6), each = nrow(X)), levels = lab_of(top6)),
  expr = as.vector(X[, top6]),
  Condition = rep(y, times = length(top6)))

p_markers <- ggplot(box_df, aes(Condition, expr, fill = Condition)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, width = 0.6) +
  geom_jitter(width = 0.15, size = 0.8, alpha = 0.6) +
  facet_wrap(~ gene, scales = "free_y", nrow = 2) +
  scale_fill_manual(values = COND_COLS) +
  labs(y = "VST-normalised expression", x = NULL,
       title = "Top-ranked discriminative marker genes") +
  theme(legend.position = "none")
print(p_markers)
ggsave(file.path(FIGDIR, "Figure4_marker_genes.png"), p_markers,
       width = 7.5, height = 5, dpi = 300)

# Export the candidate list for Part 3 (structural bioinformatics)
write.csv(deg_strict[order(deg_strict$padj),
                     c("gene","symbol","baseMean","log2FoldChange","padj","sig")],
          file.path(RESDIR, "Table_S4_all77_DEGs_for_Part3.csv"), row.names = FALSE)
```

## Session information

```{r session}
sessionInfo()
```
